Prefix-free graphs and suffix array construction in sublinear space

### 2023-08-09 manuscript/content/01_introduction.tex
integrated reviewers' suggestions

Thanks to the link to suffix arrays, prefix-free graphs have great potential to draw from this extensive knowledge.
Using the suffix array from a prefix-free graph, we can obtain several stringological data structures which are not easily obtainable for graphs.
This feature of prefix-free graphs was demonstrated in several articles \cite{2019boucher,rossi2022moni,2022pfwg,2022maria}, where the authors implicitly used similar techniques as presented here.
We think that explicitly defining and framing the prefix-free graph as a standalone pangenomic data structure can bring several benefits:

>>>>

Thanks to the link to suffix arrays, prefix-free graphs have great potential to draw from this extensive knowledge.
Using the suffix array from a prefix-free graph, we can obtain several stringological data structures which are not easily obtainable for graphs.
This feature of prefix-free graphs was implicitly demonstrated in several articles \cite{2019boucher,rossi2022moni,2022pfwg,2022maria}, where the authors used similar techniques as presented here.
We think that explicitly defining and framing the prefix-free graph as a standalone pangenomic data structure can bring several benefits:

### 2023-08-09 manuscript/content/02_pfg.tex
removed unnecessary images
integrated reviewers' suggestions

To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
Then, for each sequence in pangenome $P$, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
If the segment's sequence was not yet observed during the scan, we add it to the set of segments and assign a unique ID.
Each time we append the corresponding ID to the path representing the original sequence.

@@

After the previous steps, the set of segments and the list of paths already represent a prefix-free graph.
However, to simplify the usage of prefix-free graphs, we recommend normalizing them.
During the normalization, we sort segments lexicographically and change their IDs to correspond to the lexicographical ranks.
This relabeling is then also propagated to paths accordingly.

To illustrate the entire procedure, consider a set of sequences $P = \{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
After the partitioning, we obtain a set of segments with IDs $\{$\texttt{0:CAC}, \texttt{1:ACG}, \texttt{2:CGTAC}, \texttt{3:ACT..}, \texttt{4:ACAC}, \texttt{5:CGAC}$\}$ and a list of paths \texttt{[[0,1,2,3], [0,4,3], [0,1,5,3]]}.
After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
    \caption{
        Prefix-free graph of the running example after normalization and its representation in GFA format.
        Link lines omitted for brevity.
    }
    \label{fig:gfa}
\end{figure}

>>>>

To create a prefix-free graph from a given pangenome, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
Then, for each sequence in the pangenome, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
If the segment's sequence was not yet observed during the scan, we add it to the set of segments and assign a unique ID.
Each time we append the corresponding ID to the path representing the original sequence.

@@

After the previous steps, the set of segments and the list of paths already represent a prefix-free graph.
However, to simplify the usage of prefix-free graphs, we recommend normalizing them.
During the normalization, we sort segments lexicographically and change their IDs to correspond to the lexicographical ranks.
This relabeling is then also propagated to paths accordingly.
All construction steps can be performed in space proportional to the sum of segment lengths and the sum of path lengths, which is expected to be significantly smaller than the length of the pangenome.

To illustrate the entire procedure, consider a set of sequences $\{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
After the partitioning, we obtain a set of segments with IDs $\{$\texttt{0:CAC}, \texttt{1:ACG}, \texttt{2:CGTAC}, \texttt{3:ACT..}, \texttt{4:ACAC}, \texttt{5:CGAC}$\}$ and a list of paths \texttt{[[0,1,2,3], [0,4,3], [0,1,5,3]]}.
After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
    \caption{
        Prefix-free graph of the running example after normalization and its representation in GFA format.
        Link lines omitted for brevity.
    }
    \label{fig:gfa}
\end{figure}

### 2023-08-09 manuscript/content/03_sa_construction.tex
moved content to content
integrated reviewers' suggestions

Next, we calculate the segment join's suffix array and the longest common prefix (LCP) array.
For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
We note that these linear complexities are proportional to the length of the segment join, which is usually much smaller than the original pangenome.

@@

To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
The resulting suffix array, LCP array, segment ID array and segment positions array are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
        Segment join of segments from the running example, corresponding segment IDs, and positions.
        Permuting IDs and positions according to the inverse permutation of the suffix array results in the segment ID array and the segment position array shown in Table \ref{tab:suffix}.
    }
    \label{fig:ids_and_positions}
\end{figure}

\input{sa_table.tex}

\input{segment_table.tex}

In the suffix table, one row can represent multiple positions of the pangenome.
To identify these positions, we store some additional information in a \emph{segment table}.
For each segment, we store its length, starting positions in the pangenome and ranks of the right contexts of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\texttt{\#}$ and sentinel $\texttt{\$}$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Then, starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/path_join.png}
    \caption{
        Path join of paths from the running example and the corresponding starts and ranks. The pangenome and segments are shown above to clarify the meaning of the start values.
    }
    \label{fig:path_join}
\end{figure}

\subsection{Iteration}
With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array values.

Each row in the suffix table represents a suffix of a particular segment.
There are four cases of what the first position of these suffixes can represent within the segments:
\begin{itemize}
    \item the sentinel
    \item a separator
    \item a position within the last $k$ characters of a segment
    \item a position outside the last $k$ characters, separator and sentinel
\end{itemize}

Since the original pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.

@@

As an example, consider the row $13$ in Table \ref{tab:suffix}, the first row yielding a SA value.
Its segment ID is $0$, and from Table \ref{tab:segment}, we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

>>>>

Next, we calculate the segment join's suffix array and the longest common prefix array \cite{landau2001linear,manzini2004two}.
For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
We note that these linear complexities are proportional to the length of the segment join, which is usually much smaller than the original pangenome.

@@

To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the ISA to these arrays changes the order of computed values in correspondence to the sorted suffixes.
The resulting suffix array $\texttt{SA}$, longest common prefix array $\texttt{LCP}$, segment ID array $\texttt{ID}$ and segment positions array $\texttt{pos}$ are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
    Segment join of segments from the running example, corresponding segment IDs, and positions. Permuting IDs and positions according to the \texttt{ISA} results in the segment ID array \texttt{ID} and the segment position array \texttt{pos} shown in Table \ref{tab:suffix}. \label{fig:ids_and_positions}
    }
\end{figure}

\input{content/table_suffix.tex}

\input{content/table_segment.tex}

In the suffix table, one row can represent multiple positions of the pangenome.
% These positions can be represented as a single integer coresponding to index in the concatenated representation of a pangenome.
To identify these positions, we store some additional information in a \emph{segment table}.
For each segment of the prefix-free graph, we store its length, starting positions in the concatenated pangenome and ranks of the right contexts of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\texttt{\#}$ and sentinel $\texttt{\$}$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Then, starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/path_join.png} %% TODO: change this
    \caption{
    Path join of paths from the running example and the corresponding starts and ranks. The concatenated pangenome and segments are shown above to clarify the meaning of the start values. \label{fig:path_join}
    }
\end{figure}

\subsection{Iteration}
With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array value by value.

Each row in the suffix table represents a suffix of a particular segment.
There are four cases of what the first position of these suffixes can represent within the segments:
\begin{itemize}
    \item the sentinel $\texttt{\$}$
    \item a separator $\texttt{\#}$
    \item a position within the last $k$ characters of a segment
    \item a position outside the last $k$ characters, separator and sentinel
\end{itemize}

Since the pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.

@@

As an example, consider the row $13$ in Table \ref{tab:suffix}, the first row yielding a SA value.
Its segment ID is $0$, and from Table \ref{tab:segment}, we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix of a concatenated pangenome $P[9..] = \texttt{ACACT}$.
